---
title: "Vignette "
output: html_document
date: "2024-12-05"
---

Github Repository Name - FinalProject_Group3

Pacakge - binaryClassifier


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Introduction

This vignette provides an introduction to `BinaryClassifier`, demonstrating its key features and how to use it for your analysis. `BinaryClassifier` is designed to help users perform supervised binary classification using numerical optimization.

## Installation

To install `BinaryClassifier`, you can use the following command:
```{r}
#Install from GitHub
#library(devtools)
#install_github("AU-R-Programming/FinalProject_Group3")
library(binaryClassifier)
```
our package includes the following functionalities:

Initial Values for Optimization

Bootstrap Confidence Intervals

Confusion Matrix

-Prevalence

-Accuracy

-Sensitivity

-Specificity

-False Discovery Rate

-Diagnostic Odds Ratio

## Reading the data 

This dataset is taken from canvas course page.

```{r}
bank <- read.csv("C:\\Users\\aishw\\Downloads\\bank.csv", sep=";")

head(bank)
```

## Preparing the data

```{r}
bank$age_binary <- (bank$age - min(bank$age)) / (max(bank$age) - min(bank$age))
bank$balance_binary <- ifelse(bank$balance > median(bank$balance), 1, 0)
bank$education <- as.factor(bank$education)
head(bank)
```

```{r}
predictors <- bank[, c("age", "education")]

response <- bank$balance_binary

edu <- model.matrix(~ education - 1, data = bank)

x <- cbind(1, predictors$age, edu)
y <- response

print(dim(x))
print(length(y))
```
## Usage of Our Package functions with above dataset

We have three functions included in this package , below expalins the how to use these pacakge functions with examples.


## FUNCTION-1-Logistic Regresssion


 `logistic_regression()` function, which estimates the coefficients of a logistic regression model using numerical optimization. The function maximizes the log-likelihood to estimate the model parameters.


```{r}
log_reg <- logistic_regression(x, y)
print(log_reg)
```
This function has the inputs and outputs as below:

Inputs:-

`x`: A matrix of predictors, where each row corresponds to an observation, and each column corresponds to a feature (including a column of 1s for the intercept).

`y`: A binary response vector (0 or 1) representing the outcome variable.

Output:-

A vector of estimated coefficients.



## FUNCTION-2 -Bootstrap confidence interval


`bootstrap_ci` function  computes confidence intervals for logistic regression coefficients using bootstrap resampling.

```{r}
ci <- bootstrap_ci(x, y, n_boot = 15, alpha = 0.01)
print(ci)
```
This function has the inputs and outputs as below:

Inputs:-

`X`: A numeric matrix of predictor variables.

`y`: A binary numeric vector of responses (0/1).

`n_boot`: Number of bootstrap iterations (default is 20).

`alpha`: Significance level for confidence intervals (default is 0.05).

Outputs:-

-Estimated coefficients.

-Lower and upper bounds of the confidence intervals.



## FUNCTION-3 Confusion_metrics

The `confusion_metrics` function calculates a confusion matrix and various classification metrics for evaluating binary classification models. It takes the actual binary values `(y_true)`, predicted probabilities `(y_pred)`, and an optional classification threshold as inputs and returns a list containing the confusion matrix and performance metrics.


```{r}
y_pred <- exp(x %*% log_reg) / (1 + exp(x %*% log_reg))
metrics <- compute_metrics(y, y_pred, threshold = 0.5)
print(metrics)
```

This function has the inputs and outputs as below:

Inputs:-

`y_true`: A vector of actual binary labels (0 or 1).

`y_pred`: A vector of predicted probabilities (typically the output from a logistic regression model).

threshold: A value to convert predicted probabilities into binary predictions (default is 0.5). This allows users to set a custom decision boundary to classify probabilities as 0 or 1.

Outputs:-

-Accuracy

-precesion

-Sensitivity

-Specificity

-false discovery rate

-diagnostic odds ratio

-prevalence
```
source("C:/Users/aishw/OneDrive - Auburn University/Documents/GitHub/FinalProject_Group3/R/logistic_regression.R")
```




















